---
title: "Making your research accessible: Data Visualization and Interactive Graphics"
subtitle: "Session 4 - Maps

<html>
  <div style='float:left'></div>
  <hr color='#EB811B' size=1px width=796px>
</html>"
author: "Felix Haass"
date: ""
output: 
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts]
    nature: 
      # highlightLanguage: r
      highlightStyle: monokai
      highlightLines: true
    chakra: libs/remark-latest.min.js
    

    

      # html_document:
  #   toc: true
  #   toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      # cache = T,
                      dev = "CairoPNG",
                      # bitmapType = 'cairo',
                      warning = FALSE, message = FALSE)
htmltools::tagList(rmarkdown::html_dependency_font_awesome())

```


# Maps

- Useful to visualize **spatial** data
- Again, think of comparisons, but in space!
- Goal of this session:
  - Find maps online
  - Combine spatial data with other information (GDP etc.)
  - Plot maps with ggplot2
  

*Note:* Complex topic, very brief overview only, if you would like to know more, let me know and we organize a separate course!


---

# Example


```{r, echo = F, warning = F, eval = T, units = "in", dpi = 400, fig.width=10, fig.height = 5, fig.align = "center", out.height = "110%", out.width = "110%", cache = T}

library(rgeos)
library(maptools)
library(sp)
library(countrycode)
library(broom)
library(gapminder)
library(tidyverse)

world <- readShapeSpatial("./data/shapefiles//TM_WORLD_BORDERS_SIMPL-0.3.shp")

world$continent <- countrycode(world$ISO2, "iso2c", "continent")

# filter NAs (some Islands + Taiwan)
world <- world[!is.na(world$continent), ]

# subset Africa
africa <- world[world$continent == "Africa", ] 

# get gapminder data for Africa
data("gapminder")

gapminder$ISO3 <- countrycode(gapminder$country, "country.name", "iso3c")

gapminder1962 <- gapminder[gapminder$year == 1962, ]
gapminder2007 <- gapminder[gapminder$year == 2007, ]

# Prepare and Merge Data

africa_fort <- tidy(africa, region = "ISO3")
africa_fort <- left_join(africa_fort, 
                         gapminder1962, 
                         by = c("id" = "ISO3"))

africa_fort$year <- 1962

africa_fort2007 <- tidy(africa, region = "ISO3")
africa_fort2007 <- left_join(africa_fort2007, 
                         gapminder2007, 
                         by = c("id" = "ISO3"))

africa_fort2007$year <- 2007

# bind together

africa_fort_comb <- bind_rows(africa_fort, 
                              africa_fort2007)

africa_map <- ggplot() +
  geom_map(data = africa_fort_comb, 
           aes(map_id = id, 
               fill = lifeExp), 
           map = africa_fort) + 
  geom_path(data = africa_fort_comb, 
            aes(x = long, y = lat, group = group), 
            color = "black") + 
  labs(title = "Life Expectancy in Africa, 1962 vs. 2007\n", 
       subtitle = "Data: Gapminder\n") +
  scale_fill_distiller("Life Expectancy", 
                       palette = "RdYlGn", 
                       direction = 1) +
  coord_equal() +
  facet_wrap(~year, nrow = 1) +
  theme_void() +  
  theme(panel.grid = element_blank(), 
        legend.position = "bottom",
        legend.text.align = 0) 


print(africa_map)




```

---

# Where to get maps and spatial data?

- Each map comes in a specific spatial format, the so-called "shapefile"
- Country shapefiles and subnational data can be obtained at http://gadm.org 
  - Useful for single country shapes
  - Lower-level administrative units
  - Read the downloaded file with `country_map <- readRDS("path_to_file")`
  
Many other sources of spatial data

- Conflict data
  - Uppsala Conflict Data Program GED http://ucdp.uu.se/downloads/

Generate your own!

- Measure location of interviews with a GPS device

---

# Read spatial data in R

Note: you can find the world shapefile on the course website!

```{r, echo = T, warning = F, eval = T, units = "in", dpi = 400, fig.width=10, fig.height = 5, fig.align = "center", out.height = "110%", out.width = "110%", cache = T}

library(maptools)

world <- readShapeSpatial("./data/shapefiles/TM_WORLD_BORDERS_SIMPL-0.3.shp")

plot(world)

```

---

# Manipulate and view spatial data

The `world` object from the previous slide is a special kind of data frame, a `SpatialDataFrame`. It works like most other data frames, but has a few quirks.

**Manipulating SpatialDataFrames**: `countrycode()`

```{r,echo = T, eval = T}

library(countrycode)
library(maptools)

world <- readShapeSpatial("./data/shapefiles//TM_WORLD_BORDERS_SIMPL-0.3.shp")

world$continent <- countrycode(world$ISO3, 
                               "iso3c", # input format
                               "continent") # output format

table(world$continent)

```

---

# Manipulate and view spatial data

**Manipulating SpatialDataFrames**: `subset()`

```{r,echo = T, eval = T, fig.height = 5, fig.width = 5, dpi = 300,  units = "in", fig.align="center", out.height="150", out.width = "150"}

library(countrycode)
library(maptools)

world <- readShapeSpatial("./data/shapefiles//TM_WORLD_BORDERS_SIMPL-0.3.shp")

world$continent <- countrycode(world$ISO3, 
                               "iso3c", # input format
                               "continent") # output format

africa <- subset(world, continent == "Africa")

par(mar=c(0.1, 0.1, 0.1, 0.1))
plot(africa)

```

---

# Manipulate and view spatial data

**Manipulating SpatialDataFrames**: `View()`

```{r,echo = T, eval = F}

library(countrycode)
library(maptools)

world <- readShapeSpatial("./data/shapefiles//TM_WORLD_BORDERS_SIMPL-0.3.shp")

world$continent <- countrycode(world$ISO3, 
                               "iso3c", # input format
                               "continent") # output format


View(world@data) # note the '@' as compared to other data frames

```


---

# Merge in other data  I
  
```{r, echo = T, eval = T, warning = F, message=F}
library(countrycode)
library(maptools)
library(ggplot2)
library(tidyverse)
library(gapminder)
library(broom)

world <- readShapeSpatial("./data/shapefiles//TM_WORLD_BORDERS_SIMPL-0.3.shp")

# create continent identifier
world$continent <- countrycode(world$ISO3, 
                               "iso3c", # input format
                               "continent") # output format

# subset Africa shape file
africa <- subset(world, continent == "Africa")



```


---

# Merge in other data II

...continued from previous slide. 

```{r, echo = T, eval = T, warning = F, message = F}

# get gapminder data 
data("gapminder")

# create country identifier for merging
gapminder$ISO3 <- countrycode(gapminder$country, "country.name", "iso3c")

# only year 2007
gapminder2007 <- gapminder[gapminder$year == 2007, ]

# fortify: bring dataset into shape that ggplot can understand
africa_fort <- tidy(africa, # we use the "africa" shapefile from previous slide
                       region = "ISO3") # this becomes "id" in the fortified dataset

# join in gapminder data
africa_fort <- left_join(africa_fort, 
                         gapminder2007, 
                         by = c("id" = "ISO3"))

```

---

# Plot the map!

```{r, echo = T, warning = F, eval = T, units = "in", dpi = 400, fig.width=10, fig.height = 5, fig.align = "center", out.height = "90%", out.width = "90%", cache = T}

africa_map <- ggplot() +
  geom_path(data = africa_fort,
            aes(x = long, y = lat, group = group),
            color = "black") 

plot(africa_map)

```

---

# Plot the map! (continued)


```{r, echo = T, warning = F, eval = T, units = "in", dpi = 400, fig.width=10, fig.height = 5, fig.align = "center", out.height = "90%", out.width = "90%", cache = T}

africa_map <- ggplot() +
  geom_path(data = africa_fort,
            aes(x = long, y = lat, group = group),
            color = "black")  +
  geom_map(data = africa_fort,
           aes(map_id = id,
               fill = lifeExp),
           map = africa_fort) 

plot(africa_map)


```

---

# Plot the map! (continued)


```{r, echo = T, warning = F, eval = T, units = "in", dpi = 400, fig.width=10, fig.height = 5, fig.align = "center", out.height = "90%", out.width = "90%", cache = T}

africa_map <- ggplot() +
  geom_path(data = africa_fort,
            aes(x = long, y = lat, 
                group = group),
            color = "black")  +
  geom_map(data = africa_fort,
           aes(map_id = id,
               fill = lifeExp),
           map = africa_fort) +
  
  # appearance
  labs(title = "Life Expectancy in Africa, 2007\n", 
       subtitle = "Data: Gapminder\n") +
  scale_fill_distiller("Life Expectancy", 
                       palette = "RdYlGn", 
                       direction = 1) +
  coord_equal() +
  theme_void() +  
  theme(panel.grid = element_blank(), 
        legend.position = "bottom",
        legend.text.align = 0) 

```

---

# Plot the map! (continued)

```{r, echo = F, warning = F, eval = T, units = "in", dpi = 400, fig.width=5, fig.height = 5, fig.align = "center", out.height = "85%", out.width = "85%", cache = T}

print(africa_map)

```

---

# Exercise

Build a map of

- Latin America (hint: countrycode has only "Americas" as continent. How would you filter the world shapefile to leave out the US and Canada?)

- Merge in information on GDP per capita in 2007

- Plot the map



---

# Solution

```{r, echo = T, eval = T}

library(maptools)
library(sp)
library(countrycode)
library(gapminder)
library(tidyverse)
library(broom)

world <- readShapeSpatial("./data/shapefiles//TM_WORLD_BORDERS_SIMPL-0.3.shp")

world$continent <- countrycode(world$ISO2, "iso2c", "continent")

latam <- subset(world, continent == "Americas")

# remove US and Canada and Greenland
latam <- subset(latam, NAME != "United States" & 
                  NAME != "Canada" & 
                  NAME != "Greenland")

# get gapminder data 
data("gapminder")

gapminder$ISO3 <- countrycode(gapminder$country, "country.name", "iso3c")

gapminder2007 <- gapminder[gapminder$year == 2007, ]

```

---

# Solution II

```{r, eval = T, echo = T}
# Prepare and Merge Data

latam_fort <- tidy(latam, region = "ISO3")
latam_fort <- left_join(latam_fort, 
                         gapminder2007,
                         by = c("id" = "ISO3"))
# plot
latam_map <- ggplot() +
  geom_map(data = latam_fort, 
           aes(map_id = id, fill = gdpPercap), 
           map = latam_fort) + 
  geom_path(data = latam_fort, 
            aes(x = long, y = lat, group = group), 
            color = "black") + 
  labs(title = "GDP per capita in Latin America 2007\n", 
       subtitle = "Data: Gapminder\n") +
  scale_fill_distiller("GDP per capita", 
                       palette = "RdYlGn", 
                       direction = 1) +
  coord_equal() +
  theme_void() +  
  theme(panel.grid = element_blank(), legend.position = "bottom",
        legend.text.align = 0) 

```




---

# Solution III

```{r, echo = F, warning = F, eval = T, units = "in", dpi = 400, fig.width=5, fig.height = 5, fig.align = "center", out.height = "60%", out.width = "60%",  cache = T}

print(latam_map)

```